---
title: "Drug assay scores and visualization"
output: html_notebook
---

## Paths

```{r}
setwd("/Users/jonasamar/Stabl/Drug Study/Rscripts")
OOL_path <- "/Users/jonasamar/Stabl/Drug Study/Onset of Labor csv"
drug_assay_path <- "/Users/jonasamar/Stabl/Drug Study/Drug assay csv"
```

## Libraries

```{r}
library(tidyverse)
library(tidyselect)
library(reshape2)
library(dplyr)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(paletteer)
library(tidyr)
library(janitor)
library(lme4)
library(lmerTest)
library(utils)
library(ggplot2)
library(r2glmm)
library(sjstats)
library(gridExtra)
```

## Loading Data and Models

**OOL data and models**

```{r}
# df_model
load(paste0(OOL_path, "/pen_model_curves_cytof.rda"))
# curve.classification <- df_sum
curve.classification <- read.csv(paste0(OOL_path, "/pen_classification_curves_cytof.csv"))
# penalized dataset with outcomes (DOS) ante partum
OOL_data <- read.csv(paste0(OOL_path, "/immunome_noEGA_DOS_pen_OOL.csv")) %>%   
            filter(DOS <= 0) # we only are interested in DOS <=0
# outcomes
DOS <- OOL_data$DOS
# results from univariate analysis on OOL data
spearman_cor <- read_csv(paste0(OOL_path, "/Univariate regated OOL/SpearmanCorrelationsPval.csv"))
```

**Drug Assay data and models**

```{r}
# model.scores
load(paste0(drug_assay_path, "/mixed model scores.rda"))
# mixed.model
load(paste0(drug_assay_path, "/mixed models.rda"))
# Dosereponse
load(paste0(drug_assay_path, "/dose response.rda"))
```

## Scores

Here we want to compare the drug effect on each feature to the "normal" behavior of the feature modeled in OOL study.

### Auxiliaries score functions

**drug.effect**

```{r}
# Function which return:
# +1 for activating drugs
# -1 for inhibiting drugs
# 0 for non active (or uncertain) drugs
# based on the slope of the model and its p-value

drug.effect <- function(slope, pval, slope.threshold, pval.threshold){
  if ((abs(slope) > slope.threshold) & (pval < pval.threshold)){
    return(slope/abs(slope))
  }
  else{
    return(0)
  }
}
```

**feature.normal.behavior.at.TTL**

```{r}
# Function which returns
# 1 if the model (of the OOL feature) is increasing at TTL
# -1 if the model (of the OOL feature) is decreasing at TTL
# NA if the behavior of the feature is inconclusive (missing model or too weak slope)
# based on the model (contained in a list) itself and its p-value

feature.normal.behavior.at.TTL <- function(model, pval, slope.threshold, pval.threshold, TTL){
  if (length(pval) == 0 || is.na(pval) ||pval > pval.threshold || length(model) == 0){
    return(NA)
  }
  else{
    # Here we focus un the derivative around 0 to estimate the behavior of the feature
    model <- model[[1]]
    quadratic <- (length(model) == 3)

    # Derivative of the model at TTL
    # for quadratic model
    if (quadratic){
      deriv <- 2*model[['DOS2']]*TTL + model[['DOS']]
    } 
    # for linear model
    else{
      deriv <- model[['DOS']]
    }
    
    # Returning effect coefficient
    if (abs(deriv) > slope.threshold){
      return(deriv/abs(deriv))
    }
    else{
      return(NA)
    }
  }
}
```

**create.drug.effect.data.frame**

```{r}
# Function which creates a dataframe with 
# 1 when the drug is affecting the feature the opposite direction of its normal behavior at TTL
# -1 when the drug is affecting the feature the same direction of its normal behavior at TTL
# 0 when the effect of the drug is null
# NA when behavior of the feature is null or hard to estimate (high p-value) or if we miss data in the drug assay study

create.drug.effect.data.frame <- function(drug.name,
                                          normal.behavior.pval.threshold, 
                                          normal.behavior.slope.threshold, 
                                          drug.pval.threshold, 
                                          drug.slope.threshold,
                                          TTL){
  
  # Dataframe feature ~ stimulation containing the final effect on a drug on each feature under each stimulation
  drug.effect.dataframe <- model.scores %>%
                        filter(drug == drug.name) %>%
                        rowwise() %>%
                        mutate(drug.behavior = drug.effect(Slope, 
                                                            pval, 
                                                            slope.threshold = drug.slope.threshold, 
                                                            pval.threshold = drug.pval.threshold)) %>%
                        mutate(population = str_extract(feature, "^[^_]+"),
                               reagent_stim = str_extract(feature, "(?<=_).*$")) %>%
                        select(population, drug.behavior, reagent_stim) %>%
                        pivot_wider(names_from = reagent_stim, values_from = drug.behavior)
  
  # Concatenation of all information on the feature models + normal.behavior = +1, -1 or NA
  model.info <- merge(df_models, curve.classification, by.x = "feature", by.y = "cytof") %>%
                rowwise() %>%
                mutate(normal.behavior = feature.normal.behavior.at.TTL(model,
                                                                        pval, 
                                                                        normal.behavior.slope.threshold,
                                                                        normal.behavior.pval.threshold,
                                                                        TTL))

  # Dataframe feature ~ stimulation containing the values of normal.behavior
  behavior.dataframe <- model.info %>%
                          mutate(population = str_extract(feature, "^[^_]+"),
                                 reagent_stim = str_extract(feature, "(?<=_).*$")) %>%
                          select(population, normal.behavior, reagent_stim) %>%
                          pivot_wider(names_from = reagent_stim, values_from = normal.behavior) %>%
                          # removing features not measured in both studies (drugs and OOL)
                          # `STAT3_LPS`, `STAT5_LPS`, `STAT1_LPS`, `STAT6_LPS`, `STAT1_GMCSF`
                          # "STAT1_unstim"
                          filter(population %in% drug.effect.dataframe$population) %>%
                          select(one_of(colnames(drug.effect.dataframe)))
                        
  # Dataframe feature ~ stimulation containing the product of behavior.matrix and drug.effect.matrix
  drug.final.effect.dataframe<- merge(pivot_longer(select(behavior.dataframe, 
                                                           one_of(colnames(drug.effect.dataframe))), 
                                                   cols = -population, 
                                                   names_to = "reagent_stim", 
                                                   values_to = "normal.behavior"),
                                      pivot_longer(select(drug.effect.dataframe,
                                                          one_of(colnames(behavior.dataframe))), 
                                                   cols = -population, 
                                                   names_to = "reagent_stim", 
                                                   values_to = "drug.behavior"),
                                      by = c("population", "reagent_stim")) %>%
                                rowwise() %>%
                                mutate(drug.final.effect = ifelse(is.na(normal.behavior) ||
                                                                    is.na(drug.behavior),
                                                                  NA,
                                                                  -normal.behavior*drug.behavior)) %>%
                                select(population, reagent_stim, drug.final.effect) %>%
                                pivot_wider(names_from = reagent_stim, values_from = drug.final.effect)

  return(drug.final.effect.dataframe)
}
```

### Features of interest

For now, we don't have a ML model with selected features helping us discriminate which features are more important to look at so we are looking at two sets of features :
- the top 15 Cytof features from the OOL paper
- the features from the univariate analysis with an absolute Spearman score > 0.3

**Top 15 features from OOL paper**

```{r}
# Top immunome features from the OOL study
# we associate a coeff +1 (increasing) or -1 (decreasing) to the features
OOL.top.immunome.features <- list("CD56loCD16posNK_STAT1_IFNa"=1,
                                  # Granulocytes missing in the drug assay study
                                  "CD56hiCD16negNK_STAT1_IFNa"=1,
                                  "CD4Tnaive_MAPKAPK2_IFNa"=1,
                                  "ncMCs_CREB_GMCSF"=-1,
                                  "CD8Tcm_MAPKAPK2_unstim"=1, #-1 at TTL = 0
                                  "CD8Tem_MAPKAPK2_unstim"=1, #-1 at TTL = 0
                                  "pDCs_STAT1_IFNa"=1,
                                  "Bcells_MAPKAPK2_LPS"=1,
                                  "CD4Tem_MAPKAPK2_unstim"=1, #-1 at TTL = 0
                                  "CD8Tcm_MAPKAPK2_IFNa"=1, 
                                  "CD8Tem_MAPKAPK2_IFNa"=1,
                                  # Bcells missing in the drug assay study
                                  "CD4Tem_NFkB_IL246"=-1,
                                  "CD4Tcm_IkB_unstim"=1, #-1 at TTL = 0
                                  "mDCs_STAT6_IFNa"=1, #-1 at TTL = 0
                                  "pDCs_STAT6_IFNa"=1, #-1 at TTL = 0
                                  "mDCs_MAPKAPK2_unstim"=-1, 
                                  "pDCs_MAPKAPK2_unstim"=-1
                                  )
```

**Top features from the univariate analysis**

```{r}
# Top immunome features from the univariate analysis on OOL data
# we associate a coeff +1 (increasing) or -1 (decreasing) to the features

univariate.OOL.features <- apply(spearman_cor[abs(spearman_cor$`Spearman corr`) > 0.3,], # taking features with |Spearmanr| > 0.3
                                 1,
                                 function(row){ # associating the feature name to +1 (increasing) or -1 (decreasing)
                                   key <- row[[1]]
                                   if (!is.na(key)) {
                                     value <- as.numeric(row[[2]])/abs(as.numeric(row[[2]]))
                                     setNames(list(value), key)
                                   } 
                                 }) %>%
                                Filter(function(x) !is.null(x), .) %>%
                                unlist(recursive=FALSE)
```

### Drugs ranking

Here we rank the drugs based on their effect on the chosen features of interest. We associate a +1 score when the drug has an effect which is opposite to the "normal" behavior/ trend of the feature and -1 when the effect of the drug is colinear to the "normal" feature behavior.

```{r}
# Chosing features of interest
feature_choice <- "top_OOL_features"
OOL_features <- OOL.top.immunome.features

# Chosing thresholds over or under which we consider the effect of the drug as relevant
slope.threshold <- 0
pval.threshold <- 1e-1
```

Calculation of the scores and ranking of the features.

**drug.ranking :** dataframe containing the name of the drugs and their score.

```{r}
drug.ranking <- data.frame(drug = character(),
                           score = numeric(),
                           stringsAsFactors = FALSE)
for (drug in unique(mixed.model$drug)){
  score <- 0
  # We sum the scores over all the features for each drug
  for (feature in names(OOL_features)){
    # Getting the drug effect on the feature
    slope <- model.scores[(model.scores$feature == feature) & (model.scores$drug == drug),]$Slope
    pval <- model.scores[(model.scores$feature == feature) & (model.scores$drug == drug),]$pval
    drug_effect <- ifelse(length(slope)*length(pval) == 0, # sometimes the values of slope and/or pval is empty
                          0,
                          drug.effect(slope, pval, slope.threshold, pval.threshold))
    # Adding the individual scores of the drug each feature 
    score <- score - drug_effect*OOL_features[[feature]]
  }
  # Adding the drug and its score to drug.ranking
  drug.ranking <- rbind(drug.ranking,
                        data.frame(drug=drug,
                                   score=score,
                                   stringsAsFactors = FALSE))
}
# Arranging the drugs to see the best drugs on top
drug.ranking <- drug.ranking %>% arrange(desc(score))
drug.ranking
```

## Visualization

### Mixed models versus "normal" behavior

Plotting the mixed models of the drug on the chosen features versus the normal behavior of the OOL feature in order to see if the trends are actually opposites (or colinear) and that the models fit the data.

```{r}
# Selecting a specific feature and plotting the models for all the drugs
target_drug <- "Benzylpenicillin"

plots <- list()
i <- 0
for (target_feature in names(OOL_features)) {
  target_model <- mixed.model$model[mixed.model$feature == target_feature & mixed.model$drug == target_drug]
  pval <- model.scores[(model.scores$feature == target_feature) & (model.scores$drug == target_drug),]$pval
  
  # We only plot features on which the score has been calculated (with mixed model having a pvalue < pval < pval.threshold)
  if ((length(pval) > 0) && 
      (pval < pval.threshold) && 
      (length(target_model)>0)){
    # Counting the number of plots
    i <- i + 2
    # Target model is either an empty list or a list containing the model
    target_model <- target_model[[1]]
    
    ### PLOT THE MIXED LINEAR MODEL
    # Coefficients of the linear mixed model
    model_coefs <- coef(target_model)$ID %>% 
                      rename(Intercept = `(Intercept)`, Slope = dose) %>% 
                      rownames_to_column("ID")
    # Subset of the data
    data <- Doseresponse %>% 
                filter(feature == !!target_feature) %>%
                filter(drug == !!target_drug) %>%
                select(ID, dose, value)
    data_rani <- left_join(data, model_coefs, by = "ID")
    # Plot
    model_coef_plot <- ggplot(data = data_rani,
                              mapping = aes(x = dose,
                                            y = value, 
                                            colour = ID)) +
      geom_point(na.rm = T, alpha = 0.5, size=1) +
      geom_abline(aes(intercept = Intercept, slope = Slope, colour = ID)) +
      guides(colour = "none") +  # Remove the color legend
      ggtitle(paste0(paste0(target_feature,"\np = "),pval)) +
      xlab(NULL) +  # Remove x-axis label
      ylab(NULL) +  # Remove y-axis label
      theme(plot.title = element_text(size = 6))  # Adjust plot title size
    # Adding plot to the list of plots
    plots[[target_feature]] <- model_coef_plot

    
    ### PLOT THE ASSOCIATED OOL FEATURE
    # Getting the coefficients and class of the model
    coefs <- df_models[df_models$feature == target_feature,]$model[[1]][[1]]
    
    # Building the predictions
    y <- OOL_data[, target_feature]
    intercept <- coefs[['(Intercept)']]
    if (length(coefs) == 3){ # quadratic model
  
      a <- coefs[['DOS2']]
      b <- coefs[['DOS']]
      y_pred <- intercept + b * DOS + a * DOS^2
    }
    else{ # length(coefs) == 2, linear model
      slope <- coefs[['DOS']]
      y_pred <- intercept + slope * DOS
    }
    df <- data.frame(x = DOS, y = y, y_pred = y_pred)
    
    # Getting pvalue of the model
    pval <- curve.classification[curve.classification$cytof == target_feature,]$pval
    # Plot
    plot <- ggplot(df, aes(x = x, y = y)) +
      geom_point(color = "blue", size = 1) +
      geom_line(aes(y = y_pred), color = "red") +
      ggtitle(paste(target_feature, pval, sep=",\np = ")) +
      xlab(NULL) +  # Remove x-axis label
      ylab(NULL) +  # Remove y-axis label
      theme(plot.title = element_text(size = 6))
    # Adding plot to the list of plots
    plots[[paste0(target_feature,".OOL")]] <- plot
  }
}

# Saving the plots
pdf(paste(drug_assay_path, "/", target_drug, "_on_", feature_choice,".pdf", sep = ""))
for (j in 1:ceiling(i/16)){
  grid.arrange(grobs = plots[(16*(j-1)+1):min(16*j,i)], nrow = 4, ncol = 4)
}
dev.off()
```

### Histograms

>>> !!! Not so sure it is relevant !!!

```{r}
# Selecting a specific feature and plotting the histograms of the values/dose
target_feature <- "CD56loCD16posNK_STAT1_IFNa"

histo.plots <- list()
for (target_drug in unique(mixed.model$drug)) {
  # Subset of the data regarding the target_drug and target_feature
  data <- Doseresponse %>% 
              filter(feature == !!target_feature) %>%
              filter(drug == !!target_drug) %>%
              select(ID, dose, value)
  
  # Modifying the data by dividing each value by the mean of the patients values
  mean_values <- data %>%
                  group_by(ID) %>%
                  summarize(mean_value = mean(value))
  data_modified <- data %>%
                  left_join(mean_values, by = "ID") %>%
                  mutate(value = replace_na(value, 0)) %>%
                  mutate(value = value / mean_value) %>%
                  select(-mean_value)
  
  # Plot
  histogram_plot <- data_modified %>%
    ggplot(aes(x = dose, y=value, fill = ID)) + 
    geom_col(color="white", na.rm=TRUE) +
    guides(fill = "none") +  # Remove the color legend
    ggtitle(target_drug) +
    xlab(NULL) +  # Remove x-axis label
    ylab(NULL) +  # Remove y-axis label
    theme(plot.title = element_text(size = 12))
  # Adding plot to the list of plots
  histo.plots[[target_drug]] <- histogram_plot
}

# Printing the plots
suppressWarnings(grid.arrange(grobs = histo.plots, nrow = 3, ncol = 5))
```

### Global visualization : heatmap

```{r}
# Function to plot the pheatmap of the final drug effect
plot_pheatmap <- function(drug.name, drug.final.effect.dataframe){
  my_colors <- c("blue", "white", "red")

  data <- as.matrix(drug.final.effect.dataframe[, -1])
  rownames(data) <- drug.final.effect.dataframe$population
  
  data[is.na(data)] <- 0
  
  pheatmap(data, 
           color = my_colors,
           na_color = "grey",
           cluster_rows = TRUE,
           cluster_cols = TRUE,
           main = drug.name,
           cellwidth = 10, 
           cellheight = 10,
           legend_breaks = c(-1, 0, 1),
           legend_labels = c("Drug effect colinear\nto normal pregnancy", 
                             "Missing Value\nor\nNot significant", 
                             "Drug effect opposite\nto normal pregnancy")
           )
}
```

Saving the heat maps for a chosen set of thresholds.

```{r}
# Parameters for the heat map scores
TTL <- -84
normal.behavior.pval.threshold <- 5e-2
normal.behavior.slope.threshold <- 1e-5
drug.pval.threshold <- 5e-2
drug.slope.threshold <- 0.

# Plotting all the heat maps
pdf(paste(drug_assay_path, "/Drug effect heatmaps TTL=",TTL, ".pdf", sep=""), width = 12, height = 6)
for (target_drug in unique(mixed.model$drug)){
  plot_pheatmap(target_drug, 
                create.drug.effect.data.frame(
                  target_drug,
                  normal.behavior.pval.threshold, 
                  normal.behavior.slope.threshold, 
                  drug.pval.threshold, 
                  drug.slope.threshold,
                  TTL
                ))
}
dev.off()
```